---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Evan Lim"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#Loading packages
library(tidyverse)
library(lubridate)
library(agricolae)
library(here)
library(corrplot)

#Checking WD
getwd()
here()

#Importing data and correcting dates
Lake <- read.csv(here("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                 stringsAsFactors = TRUE)
          

Lake$sampledate <- mdy(Lake$sampledate)

#2
#Building and setting theme
a7theme <- theme_classic(base_size = 14)+
  theme(
    axis.title = element_text(color = "darkblue"),
    panel.background = element_rect(fill = "gray"),
    legend.position = "top")

theme_set(a7theme)
  
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across
all lakes
Ha: Mean lake temperature recorded during July does change with depth across all
lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Lake.wrangled <- Lake %>% 
  mutate(Month = month(sampledate)) %>% 
  filter(Month == "7") %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()
  

#5
tempvsdepth <- ggplot(Lake.wrangled, 
                      aes(x=depth, 
                          y=temperature_C)) +
                  geom_point() +
                  geom_smooth(method = lm) +
                  ylim(0,35) +
                  ylab("Temperature (C)") +
                  xlab("Lake Depth")
              
print(tempvsdepth)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests a negative correlation between lake depth and
temperature. So, as lake depth increases, temperature goes down. However,
there is a fairly wide distribution of points, especially at smaller depth values.
This distribution does taper off after depth values of 8 or so, and there is
something of a plateau, suggesting lake temps are consistently low past a certain
depth.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

tempvsdepth.regression <- lm(data = Lake.wrangled, temperature_C~depth)

summary(tempvsdepth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model reports that there is indeed a statistically significant,
negative correlation between lake temperature and depth. According to the model, 
73.87% of the variability in temperature is explained by changes in depth, as 
indicated by the adjusted r-squared value of 0.7387. This is based on 9726 degrees 
of freedom with a statistical significance of 2.2e-16, which is well below .05. 
For every 1m change in depth, temperature is predicted to drop by -1.95 degrees
celsius. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
lakeAIC <- lm(data = Lake.wrangled, 
              temperature_C~depth + 
                year4 + 
                daynum)
step(lakeAIC) 


#10
lake.bestregression <- lm(data = Lake.wrangled, 
                          temperature_C~depth + 
                           year4 + 
                           daynum)
summary(lake.bestregression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggested that we include depth, year, and day number.
Removing any one of these values resulted in a higher AIC value, indicating a
worse fit. The multiple regression for all those variables reports that 74.12%
of the variance in temperature can be explained by the three variables. This is
only a slight improvement over using only depth as an explanatory variable, 
which was responsible for 73.87% of the variance. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

summary(Lake.wrangled)
summary(Lake.wrangled$lakename)

#Running anova
Lake.wrangled.anova <- aov(data = Lake.wrangled, temperature_C ~ lakename)
summary(Lake.wrangled.anova)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The ANOVA test has resulted in a p value less than .05, meaning we
reject the null hypothesis of the mean temperature among all lakes being the same.
In other words, there is a statistically significant difference in mean temps
among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
tempvsdepth.bylake <- ggplot(Lake.wrangled, 
                      aes(x=depth, 
                          y=temperature_C,
                          color=lakename)) +
                  geom_point(alpha = 0.5) +
                  geom_smooth(method = lm) +
                  ylim(0,35) +
                  ylab("Temperature (C)") +
                  xlab("Lake Depth") +
                  labs(color = "Lake Name")+
                  theme(legend.position = "right")

print(tempvsdepth.bylake)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lake.wrangled.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Central Long Lake (-3.85), Crampton Lake (-2.03), East Long Lake (3.04), Hummingbird Lake (2.54), Tuesday Lake (-2.24), West Long Lake (-1.73)

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
Crampton.Ward <- Lake.wrangled %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))


```

>Answer: 
