---
title: "Assignment 5: Data Visualization"
author: "Evan Lim"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Workspace set up, loading packages, checking WD and importing files
library(tidyverse)
library(here)
library(cowplot)

# Checking WD
getwd()
here()

# Importing datasets

Peter.Paul <- read.csv(here("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

Neonics <- read.csv(here("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

#2 Changing formats to date
Peter.Paul$sampledate <- ymd(Peter.Paul$sampledate)
Neonics$collectDate <- ymd(Neonics$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Making a custom theme

eggyolk_theme <- theme_base() +
  theme(
    axis.title = element_text(color = "orange"),
    legend.position = "right",
    plot.title.position = "plot",
    plot.background = element_rect(color = "gray")
  )

theme_set(eggyolk_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 Plotting total phosphorus by phosphate for Peter and Paul Lakes

Peter.Paul %>%  
  ggplot(
    aes(
      x = tp_ug,
      y = po4,
      color = lakename)
    ) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
) +
  xlim(0,100)+
  ylim(0,45)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
Peter.Paul$month <- factor(
  Peter.Paul$month,
  levels = c("5","6","7","8","9","10","11"),
  labels = c("May","Jun","Jul","Aug","Sep","Oct","Nov")
)
 
unique(Peter.Paul$month)

tempplot <- ggplot(Peter.Paul,
    aes(
      x = month,
      y = temperature_C,
      color = lakename)) +
    geom_boxplot()+
  theme(
    axis.title.x = element_blank(),
    legend.position = "none",
    ylab("Temp")
    )

  
TPplot <- ggplot(Peter.Paul,
    aes(
      x = month,
      y = tp_ug,
      color = lakename)) +
    geom_boxplot()+
   theme(
    axis.title.x = element_blank(),
    ylab("TP")
    )
  

TNplot <- ggplot(Peter.Paul,
    aes(
      x= month,
      y = tn_ug,
      color = lakename)) +
    geom_boxplot()+
   theme(
     legend.position = "none",
     ylab("TN")
         )
  

plot_grid(tempplot, TPplot, TNplot, nrow = 3, align = "v")

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: